\documentclass[12pt,a4paper]{article}
\usepackage{fontspec}
\defaultfontfeatures{Mapping=tex-text}
\usepackage{xunicode}
\usepackage{xltxtra}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{bbold}
\usepackage{enumitem}
\usepackage{graphicx}
\usepackage[left=1.5cm,right=1.5cm,top=1.5cm,bottom=1.5cm]{geometry}
\usepackage{nicefrac}
\usepackage{xspace}

\providecommand{\id}{\mathbb{1}}
\providecommand{\gcdd}{\mathrm{gcd}}
\providecommand{\lcm}{\mathrm{lcm}}
\providecommand{\FAG}{finite abelian group\xspace}
\providecommand{\bemph}[1]{\emph{\textbf{#1}}}
\providecommand{\abs}[1]{\left \vert #1 \right \vert}
\providecommand{\index}[2]{\left[ #1 \,:\, #2 \right] #1 \right \vert}
\providecommand{\hint}{\emph{Hint: }}
\providecommand{\soln}{\textbf{Solution: }}
\providecommand{\set}[1]{\left \lbrace #1 \right \rbrace}
\providecommand{\st}{such that\xspace}
\providecommand{\JK}{Joel Klassen\xspace}
\providecommand{\CV}{Christophe Vuillot\xspace}
\providecommand{\JH}{Jonas Helsen\xspace}
\providecommand{\LB}{Lennart Bittel\xspace}
\renewcommand{\iff}{\textrm{if and only if}\xspace}
\providecommand{\inv}{^{-1}}
\providecommand{\gen}[1]{\left \langle #1 \right \rangle}
\providecommand{\hism}{homomorphism\xspace}
\providecommand{\iism}{isomorphism\xspace}
\providecommand{\mod}{\textrm{ mod }}

\title{An Introduction to the Theory of Groups}
\author{Joseph J. Rotman}

\begin{document}
\maketitle
Fourth Edition
\section*{Problems}
\subsection*{Chapter 1}
\paragraph*{1.13}

\begin{enumerate}[label=(\roman*)]
\item A permutation $\alpha \in S_n$ is \bemph{regular} if either $\alpha$ has no fixed points and it is the product of disjoint cycles of the same length, or $\alpha = \id$. 
Prove that $\alpha$ is regular iff it is a power of an $n$-cycle $\beta$; that is, $\alpha = \beta^m$ for some $m$. (\hint if $\alpha = (a_1 a_2 \ldots a_k)(b_1 b_2 \ldots b_k)\ldots(z_1 z_2 \ldots z_k)$, where there are $m$ letters $a$, $b$, \ldots, $z$, then let $\beta = (a_1 b_1 \ldots z_1 a_2 b_2 \ldots z_2 \ldots a_k b_k \ldots z_k)$.)

\soln $\beta^m$ takes $a_1$ through $b_1 \ldots z_1$ to $a_2$ as desired. 
$\id$ can be expressed as $\beta^n$ for $\beta(j) = j+1$ an $n$-cycle. 
For a general regular $\alpha$, disjointess of the sets $a_j$, $b_j$, \ldots $z_j$ guaranteed that the $\beta$ from the hint is an $n$-cycle. 
If there's some $n$-cycle $\beta$ with $n=mk$, and we take the $m$th power, we also get $m$ disjoint, length-$k$ cycles, as desired.
\item If $\alpha$ is an $n$-cycle, then $\alpha^k$ is a product of $\gcd(n, k)$ disjoint cycles, each of length $\nicefrac{n}{\gcd(n, k)}$.

\soln $\alpha^n = \id$. If $n$ is a multiple of $k$, then $\alpha^n = \left( \alpha^{k} \right)^{\nicefrac{n}{k}}$.
$\alpha^k$ would then be a product of $k$ $\nicefrac{n}{k}$-cycles.
In the case where $n$ is not a multiple of $k$, but they have a non-trivial gcd, then starting at $\alpha_0$, $\alpha$ would take us to $\alpha_1$. 
$\alpha^k$ will take us to $\alpha_k$. 
It takes $\alpha_k$ to $\alpha_{2k}$, and so on until we get to $\alpha_{mk}$ = $\alpha_0$. 
This happens if $m = \frac{n}{\gcd(n,k)}$, but I don't know how to prove that. 
\item If $p$ is prime, then every power of a $p$ cycle is either a $p$-cycle or $\id$.

\soln This is a corollary of the last exercise, noting that $\gcd(p,k)=1$ if $k \neq p$ and $p$ if $k = p$.
\end{enumerate}

\paragraph*{1.17}

How many $\alpha \in S_n$ are there with $\alpha^2 = \id$?

\soln There's $\id$, and there's disjoint unions of transpositions. 
In terms of single transpositions, there are ${n \choose 2}$ of them. 
If I'm going to put together a product of $j$ transpositions, there are ${n \choose 2}$ ways to choose the first transposition, ${{n-2} \choose 2}$ ways to choose the second, and ${{n-2j} \choose 2}$ ways to choose the $j$th. Since any permutation of these transpositions is equivalent, I ought to get
\begin{equation*}
1 + \sum_{j=1}^{\nicefrac{n}{2}} \dfrac{1}{j!} \prod_{k=0}^{j} {{n-2k} \choose 2}
\end{equation*}
or something, it's not important.

\paragraph*{1.26}

A group for which $x^2=\id$ for all $x$ must be Abelian.

\soln We know that $aa = aea = abba = \id$, and that $abab = \id$. 
This implies that $ab$ and $ba$ must both be equal to $b\inv a\inv$. 

\paragraph{1.27}

\begin{enumerate}[label=(\roman*)]
\item Let $G$ be a \FAG containing no elements $a \neq e$ with $a^2 = e$. 
Evaluate $a_1 \ast a_2 \ast \ldots \ast a_n$, where $a_1,\, a_2,\, \ldots, \, a_n$ is a list of all elements in $G$ with no repetitions.

\soln Just for laughs, let's invert this big element.
From the result of Exercise 1.23, we get $(a_1 \ast a_2 \ast \ldots \ast a_n)\inv = a_n\inv \ast a_{n-1}\inv \ast \ldots \ast a_1\inv$.
Let $A$ be another name for this big element, so I don't have to \LaTeX it all out. 
The inverses of the individual group elements are unique elements of the group themselves, so the inverse of $A$ is another product of all elements of the group. 
$G$ is abelian, so $A\inv = A$, since all permutations of products are equivalent.
This means $A^2=e$, and the only element of $G$ for which that holds is $\id$.
\item Prove \bemph{Wilson's Theorem}: If $p$ is prime, then
\begin{equation*}
(p-1)! = -1 \mod p.
\end{equation*}
(\hint The nonzero elements of $\mathbb{Z}_p$ form a multiplicative group.)

\soln As far as I'm concerned, this completely contradicts the last exercise, since $-1$ is not the multiplicative identity unless $p=2$. 
One interesting thing to note is that $(-1)^2 = 1 = 1^2$, violating the assumption of part (i).
Now things start to get a little clearer. The inverse is unique, so, for all numbers from $2$ up to $p-2$, the multiplicative inverse mod $p$ is \emph{also} in the set \texttt{range}(2, $p-1$) (ranges are taken to be Python-style). 
That means that $(p-2)! = \id$, and $(p-1)! = p-1$, as desired.
\end{enumerate}

\paragraph*{1.31}

Let $G$ be a group, let $a \in G$, and let $m$ and $n$ be relatively prime integers. 
If $a^m=\id$, show that there exists a $b$ \st $a=b^n$. 
(\hint There are integers $s$ and $t$ \st $sm + tn = 1$. )

\soln (Special thanks to \JK and \CV for their assistance.) We know $a^m = \id$, so that $a^{m + 1} = a$. 
From the hint, $a^{m + sm + tn} = a$, and we can cancel multiples of $m$ to obtain $a^{tn} = a$, so we set $b = a^t$ and get what we were after.

\paragraph*{1.42}

Let $G=\set{x_1,\,x_2,\,\ldots,\,x_n}$ be a set equipped with an operation $\ast$, let $A=\left[ a_{ij} \right]$ be its multiplication table (i.e. $a_{ij} = x_i \ast x_j$), and assume that $G$ has a two-sided identity $e$: $e \ast x = x \ast e = x$ for all $x \in G$.   
\begin{enumerate}[label=(\roman*)]
\item Show that $\ast$ is commutative \iff $A$ is symmetric.

\soln This is true by definition.
\item Show that every element $x \in G$ has a (two-sided) inverse (i.e. there is an $x' \in G$ ) \st $x' \ast x=x \ast x' = e$. \iff $A$ is a \bemph{Latin Square} (i.e. all rows and columns are permutations of $G$). 

\soln (Thanks to \JK and \CV for basically doing this problem for me). If $A$ is a Latin Square, then there exists a left inverse and a right inverse for $x$, since $\id$ must appear in the row and column corresponding to $x$. $zx=\id$, $xy=\id$, therefore $zxy=z$, but $zx=\id$, so $y=z$.

Likewise, if $A$ is not a Latin square, then there are different elements $y$ and $z$ \st $xy=xz=w$. 
If $x$ has a left inverse, then $y=z$ and we have a contradiction. 
If it doesn't, we've proven what we want to prove.

\item Assume that $e=x_1$ so that the first row of $A$ has $a_{1i} = x_i$. 
Show that the first column of $A$ has $a_{i1} = x_i\inv$ for all $i$ \iff $a_{ii}=\id$ for all $i$.

\soln This is mad trivial.
\item With the multiplication table as shown in (iii), show that $\ast$ is associative \iff $a_{ij}a_{jk} = a_{ik}$.

\soln

\textbf{If: } The trick here is that, if the matrix is arranged such that $\id$ is on the diagonal, then $a_{ij} = x_i x_j\inv$. 
If multiplication is associatve, $a_{ij} a_{jk} = x_i x_j\inv x_j x_k\inv = x_i x_k\inv = a_{ik}$. \qed

\textbf{Only If: } Every $x_k$ can be expressed as $x_i x_j\inv$ for fixed $x_i$, since the multiplication table is a latin square. 
This implies that the product of three elements $x_k x_l x_m = x_i x_j\inv x_j x_n\inv x_n x_o\inv = a_{ij}a_{jn}a_{no}$. 
We can evaluate this product in either order, using the assumed product: $a_{ij}a_{jn}a_{no} = a_{in}a_{no} = a_{ij}a_{jo} = a_{io}$. \qed
\end{enumerate}

\paragraph*{2.3}

The set-theoretic union of two subgroups is a subgroup \iff one is contained in the other. 
Is this true if we use three subgroups?

\soln No, see Bruckheimer et al, 1970: \texttt{https://www.jstor.org/stable/pdf/2316854.pdf}.

\paragraph*{2.4}

Let $S$ be a proper subgroup of $G$. 
If $G - S$ is the complement of $S$, prove that $\gen{G - S} = G$.

\soln We know that $\gen{G - S}$ contains $G - S$, so all we have to prove is that the elements of $S$ can be generated by multiplying together two things in $G - S$.
Pick an element of $G - S$ $g$. 
$g^{-1} \notin S$, since that would contradict the inclusion of the inverse. 
Also, $sg^{-1} \notin S$ for any $s \in S$, since that would contradict closure.  
However, $sg^{-1} \cdot g = s$, so we can find two elements of $G - S$ that generate $s$ under multiplication. 

\paragraph*{2.5}

Let $f: \, G \rightarrow H$ and $g: \, G \rightarrow H$ be \hism{s}, and let
\begin{equation*}
K = \set{a \in G: \, f(a) = g(a)}.
\end{equation*}
Must $K$ be a subgroup of $G$?

\soln We need: 
\begin{itemize}
	\item the identity to be in the set.
	This is guaranteed by Theorem 1.13 -- $f(\id) = \id'$, so $\id \in K$.
	\item closure under the inverse. 
	This is also guaranteed by Theorem 1.13 -- $f(a\inv) = f(a)\inv = g(a)\inv = g(a\inv)$.
	\item closure under multiplication. 
	This is guaranteed by the defining property of the \hism -- $f(a b) = f(a) f(b) = g(a) g(b) = g(a b)$. 
\end{itemize}
$K$ is a subgroup. \qed

\paragraph*{2.7}

If $n > 2$, then $A_n$ is generated by all the $3$-cycles. (\hint $(ij)(jk)=(ijk)$ and $(ij)(kl)=(ijk)(jkl)$).

\soln Parity is defined as the number of transpositions in a decomposition of a permutation.
For each adjacent pair of transpositions in such a decomposition, we can apply one of the two formulae in the hint to express it as a product of 3-cycles.

\paragraph*{2.8}

Imbed $S_n$ as a subgroup of $A_{n+2}$, but show that, for $n \geq 2$, $S_n$ cannot be imbedded in $A_{n+1}$.

\soln The first part is trivial. 
If we have elements $n + 1$ and $n + 2$, we can decide whether to multiply a permutation by $(n + 1 \, n + 2)$ on its way into the subgroup, and everything becomes even. 
The second part is tricky. 
We can easily prove that $S_2$ can't be imbedded in $A_{3}$, since $A_3$ is generated by the 3-cycle $(1\,2\,3)$ which is order 3 (as are all of its subgroups), and $S_2$ is order 2.
\JH says that, in order to be imbedded, the order of the large group has to be an integer multiple of the order of the small group. 
$\abs{S_n} = n!$ and $\abs{A_{n + 1}} = \nicefrac{(n + 1)!}{2}$, so the ratio is $\nicefrac{n+1}{2}$, which is not an integer if $n$ is even. 
We could also (try to) use the fact that, for $n + 1 \geq 5$, $A_{n + 1}$ is simple, and $S_n$ is not, since it has $A_n$ as a normal subgroup.
This would leave us with the $n=3$ case, which we can allegedly show by seeing that $A_4$ has no subgroups of order 6.
I'm not quite happy with this solution.

\paragraph*{2.9}

\begin{enumerate}[label=(\roman*)]
\item Prove that $S_n$ can be generated by $\set{(1 \, k), k \in \mathbb{Z}_n}$

\soln We know that every element of $S_n$ can be expressed as a product of transpositions, and $(j \, k) = (1 \, k) (1 \, j) (1 \, k)$, so we're good. \qed
\item Prove that $S_n$ can be generated by $\set{(j \, j + 1 \mod n), j \in \mathbb{Z}_n}$. 

\soln If we can generate the generators from the generating set in the last part, we're good. 
Let's use recursion, because this is \emph{mathematics}. $(1 \, k) = (1 \, k-1) (k - 1 \, k) (1 \, k - 1)$.
\item Prove that $S_n$ can be generated by the two elements $(1 \, 2)$ and $C=( 1 \, 2 \, \cdots \, n)$.

\soln First, if I take the $n$th power of $C$, I get $\id$.
This means that $C\inv$ is the $n-1$th power, and I can use it to generate other stuff.
Let's look at $C^k(1 \, 2)C^{-k}$. 
This operation takes $k + 1 \rightarrow 1 \rightarrow 2 \rightarrow k + 2$ and $k + 2 \rightarrow 2 \rightarrow 1 \rightarrow k + 1$, so I can generate the generators from the generating set in the last part.
\item Prove that $S_4$ is not generated by $A=(1 \, 3)$ and $C=(1 \, 2 \, 3 \, 4)$.

\soln I can write down an arbitrary product of these operators as $C^{i_0}AC^{i_1}AC^{i_2}\ldots A C^{i_m}$.
Knowing that $C\inv$ is generated, I express the product as
\begin{equation*}
C^{i_0}AC^{-i_0} C^{i_0 + i_1}A C^{-(i_0 + i_1)} \ldots C^{\sum_{j=0}^{m-1} i_j } A C^{-\sum_{j=0}^{m-1} i_j } C^{-\sum_{j=0}^{m} i_j }
\end{equation*}
The only two values of $C^i A C^{-i}$ are $(1 \, 3)$ and $(2 \, 4)$, and the only 3 non-$\id$ values of $C^i$ are $(1 \, 2\, 3\, 4)$, $(1 \, 3) (2 \, 4)$, and $(1 \, 4\, 3\, 2)$. 
Brute force enumeration of the words of this set reveals that $(1 \, 2)$ isn't generated. \qed
\end{enumerate}
\paragraph*{2.12}

\begin{enumerate}[label=(\roman*)]
\item Prove that every group $G$ of order 4 is isomorphic either to $\mathbb{Z}_4$ or the 4-group $\mathbf{V}$.

\soln Without loss of generality, the elements of a group of order 4 are $\id$, $a$, $b$, and $ab$. 
Each of these elements needs an inverse, and the inverse of the inverse is the element itself, so that if $a\inv = b$, then $b\inv = a$ also. 
Up to permutation of labels, there are two possible scenarios, either $a\inv = a$, $b\inv = b$, or $a\inv = b$. Writing out the multiplication tables then either gives us $\mathbb{Z}_4$ or $\mathbf{V}$.


\item If $G$ is a group with $\abs{G} \leq 5$, then $G$ is abelian.

\soln Order 2 is trivial, there's only one non-trivial group element.
3 is prime, so the group has to be cyclic, and therefore abelian.
Order 4 only permits the two groups we saw earlier, and 5 is prime.
\end{enumerate}

\paragraph*{2.16}

If $H \leq G$ has index 2, then $a^2 \in H$ for every $a \in G$

\soln (\CV knocked this out in 60 seconds.) If $a$ is in $H$, we're done. 
If not, it can be expressed as $gh$ for some $h$ in $H$. 
Let's assume that the square is not in the subgroup. 
This implies $ghgh = gh'$, $\therefore hgh=h'$, $\therefore gh = h\inv h'$, which is a contradiction, since $h\inv h'$ is in $H$. 

\paragraph*{2.17}
\begin{enumerate}[label=(\roman*)]
\item If $a ,\, b \in G$ commute and if $a^m = \id = b^n$, then $(ab)^k = \id$, where $k=\lcm(m,n)$. 

\soln They commute so $(ab)^k = a^k b^k$. 
\item Let $A$ and $B$ be
\begin{equation*}
A = \begin{bmatrix} 0 & -1 \\  1 &  0 \end{bmatrix}, \qquad
B = \begin{bmatrix} 0 &  1 \\ -1 & -1 \end{bmatrix}
\end{equation*}
Show that $A^4 = B^3 = \id$, but $AB$ has infinite order.

\soln Let's take a look at the effect of $AB$ acting from the left on a matrix.
\begin{equation*}
\begin{bmatrix} 1 & 1 \\ 0 & 1 \end{bmatrix}
\begin{bmatrix} a & b \\ c & d \end{bmatrix}
=
\begin{bmatrix} a+c & b+d \\ c & d \end{bmatrix}
\end{equation*}
This preserves $c$ and $d$, so that if they were 0 and 1 on the way in, it'll be the same on the way out. 
\begin{equation*}
\begin{bmatrix} 1 & 1 \\ 0 & 1 \end{bmatrix}^k = \begin{bmatrix} 1 & k \\ 0 & 1 \end{bmatrix}
\end{equation*}
Note that this matrix doesn't generate the Fibonacci sequence like I thought it would, which is why I did this trivial exercise. 
\end{enumerate}

\paragraph*{Note}
There comes a point in every yellow book where the physicist reading it will begin to skip non-trivial exercises, simply because the physicist in question is uninterested in the answers to those questions. 
This is that point.

\paragraph*{2.30}

If $S \leq G$ and $\index{G}{S}=2$, then $S \lhd G$.

\soln We'd like to show that $gsg\inv \in S$. There are, just as in exercise 2.16, two cases. If $g \in S$, in which case $g\inv \in S$ and $gsg\inv $ is just a product of three subgroup elements. 
If $g \notin S$, then $g=as_0$ with $s_0 \in S$. 
Let's assume that $gsg\inv \notin S$. 
This implies that $as_0 s s_0\inv a\inv = as_1$, which implies $s_0 s s_0\inv a\inv = s_1$.
Inverting both sides, we obtain a contradiction: $a s_0 s\inv s_0 \inv$ must be in the coset, but $s_1\inv$ must be in the subgroup.

\paragraph*{2.32} 

If $H \leq G$ then $H \lhd G$ \iff for all $x, y \in G$, $xy \in H$ \iff $yx \in H$. 

\soln (Solved with the assistance of \LB) If $H \lhd G$ and $xy \in H$, then $gxyg\inv \in H$ for all $g$.
Setting $g=x\inv$ gets us $yx\in H$. 
If $xy \in H$ implies $yx \in H$, then $xy(yx)\inv = xyx\inv y\inv$ is in $H$. 
This implies that, if $y\inv \in H$, then $xyx\inv \in H$ for all $x$. 
$y\inv \in H$ also implies $y \in H $. \qed  

\paragraph*{2.38}

If $H$ and $K$ are normal subgroups of $G$, show that $H \vee K \lhd G$. 

\soln $ghkg\inv = ghg\inv gkg\inv = h'k' \in H \vee K$. \qed

\paragraph*{2.39}

Prove that if a normal subgroup $H$ of $G$ has index $n$, then $g^n \in H$ for all $g \in G$. 

\soln We (\CV and I) are going to prove that $g^n$ is not in any coset of $H$, and has to, therefore, be in $H$.
We begin by showing that, if the coset representatives are $a_j$ for the $j$th coset, then $a_j h_0 a_k h_1$ is not in coset $j$ or $k$. 
Let's assume that it's in each of these cosets, and derive contradictions.
\begin{flalign*}
a_j h_0 a_k h_1 &= a_j h_2 \\
h_0 a_k h_1 &= h_2 \\
a_k h_1 &= h_2 h_1\inv
\end{flalign*}
That's a contradiction, since an element in coset $k$ can't be in the subgroup. 
\begin{flalign*}
a_j h_0 a_k h_1 &= a_k h_2
\end{flalign*}
Let's use normality, 
\begin{flalign*}
a_j h_0 h_1 a_k &= h_2 a_k \\
a_j h_0 h_1 &= h_2 \\
a_j h_0 &= h_2 h_1\inv
\end{flalign*}
Again, contradiction.

Now, let's consider $g^n$. 
$g=a_1 h$ for some $a_1$ in the coset representatives, and $a_1 h a_1 h$ cannot be equal to $a_1 h'$, so we say that, if it's not in $H$, it must be equal to some $a_2 h'$ from a separate coset. 
Then $g^3 = a_1 h a_1 h a_1 h$ can't be from cosets 1 or 2, $g^4$ can't be from cosets 1 through 3, and $g^n$ can't be from any non-$H$ coset.

\paragraph*{2.41}

Let $G$ be a finite group of odd order, and let $x$ be the product of all the elements
of $G$ in some order. Prove that $x \in G'$.

\soln It feels like we need to use the fact that the group is odd-sized, but let's first remember that $\id$ is one of those elements, so the rest of the group is even-sized. 
Also, if $a \in G$, then $a\inv \in G$, so we have a randomly-arranged sequence containing each element and its inverse. 
I don't know what to do with this information, so let's try some small groups. 
Order 3 has 3 possible sequences, $aba\inv b\inv$, $aa\inv bb\inv$, $abb\inv a\inv$, two of which are the identity, and the third is a commutator. 
That's not too enlightening, though, so let's try one at order seven. 
A sequence could look like $a b c a\inv b\inv c\inv$. 
If we're quite clever, we can write this as $a (b c) a\inv (c b)\inv$, insert the identity $a (b c) a\inv (b c)\inv (b c) (c b)\inv$ and expand to $a (b c) a\inv (b c)\inv b c b\inv c\inv$, and we're done.

That's sort of cool, because it doesn't depend on the size of the group. 
I could take some arbitrary product, find the inverse of the first element somewhere in the expression, and produce an element of the commutator group, effectively removing a variable from the expression. 

But what do I do if the first element in the sequence is self-inverse? If the group is odd-order, there must be another self-inverse element, since $a\inv$ no longer exists as a distinct element. It also doesn't appear in the sequence. 
Let's consider that order-seven example again: $abcb\inv c\inv d$. 
With the same trick as before, we can move the $a$ to be next to $d$, but this leaves us with the task of showing that $ad \in G'$, which I don't think it is.

To deal with this, we note that there's an element of the group which is equal to $da$, so we express it that way in the sequence. 
Then, we use the commutator trick to move $a$ and $d$ next to $da$.
The sub-sequence this produces is either $aadd = \id$ or $adad$ which is a commutator. 

This can only be accomplished when the number of self-inverse elements is even, so it'll only work when the group order is odd.  
\end{document}

[label=(\roman*)]